---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ggstackplot

<!-- badges: start -->
[![R-CMD-check](https://github.com/KopfLab/ggstackplot/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/KopfLab/ggstackplot/actions/workflows/R-CMD-check.yaml)
[![Documentation](https://img.shields.io/badge/docs-online-green.svg)](https://ggstackplot.kopflab.org/)
[![codecov](https://codecov.io/gh/kopflab/ggstackplot/branch/main/graph/badge.svg?token=SN0YDIJ6Y6)](https://app.codecov.io/gh/kopflab/ggstackplot)
<!-- badges: end -->

Have you ever wanted to compare multiple panels of line plots that do not share coordinate schemes? These kinds of plots are ubiquitous in the Earth sciences, but there is not an easy way to create them with ggplot facets.

## Installation

You can install ggstackplot from [GitHub](https://github.com/) with:

```{r, eval = FALSE}
if(!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
devtools::install_github("KopfLab/ggstackplot")
```

## Show me some code

```{r example, fig.width=7, fig.height=7}
library(ggstackplot)
ggstackplot(
  mtcars,
  x = mpg,
  y = c(`weight [g]` = wt, qsec, drat, disp),
  color = c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3"),
  overlap = c(1, 0, 0.3)
)
```

## Next steps

- check out the `Features` vignette for details on all available functionality

