---
title: "ggstackplot features"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggstackplot features}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(ggstackplot)
```

# Main Arguments

## `x` and `y` arguments

### vertical stack

```{r}
# select any number of variables to make the stack
mtcars |> 
  ggstackplot(
    x = mpg, y = c(wt, qsec, drat)
  )

# the selection order translates into stack order
mtcars |> 
  ggstackplot(
    x = mpg, y = c(drat, wt, qsec)
  )

# use any valid tidyselect selection syntax
mtcars |> 
  ggstackplot(
    x = mpg, y = c(4, "carb", starts_with("d"))
  )

# use any valid tidyselect renaming syntax to rename stack panels
mtcars |> 
  ggstackplot(
    x = c(`mpg [units]` = mpg), 
    y = c(`weight [tons]` = wt, `speed` = qsec, drat)
  )
```

### horizontal stack

```{r}
# all examples shown in this document work the same way for a horizontal
# stack simplfy by switching out the x and y assignments
mtcars |> 
  ggstackplot(
    y = mpg, x = c(wt, qsec, drat)
  )
```


## `color` argument

```{r}
# select any specific colors for each plot
mtcars |> 
  ggstackplot(
    x = mpg, y = c(wt, qsec),
    color = c("#E41A1C", "#377EB8")
  )
```

```{r}
# likewise for the horizontal stack version
mtcars |> 
  ggstackplot(
    y = mpg, x = c(wt, qsec),
    color = c("#E41A1C", "#377EB8")
  )
```


## `remove_na` argument

```{r,  message=FALSE}
library(dplyr)

# default (NAs are removed so lines are not interrupted)
mtcars |> 
  add_row(mpg = 22, wt = 5, qsec = NA) |>
  ggstackplot(
    x = mpg, y = c(wt, qsec),
    color = c("#E41A1C", "#377EB8")
  )

# explicit `remove_na` = FALSE
mtcars |> 
  add_row(mpg = 22, wt = 5, qsec = NA) |>
  ggstackplot(
    x = mpg, y = c(wt, qsec),
    color = c("#E41A1C", "#377EB8"),
    remove_na = FALSE
  )
```

## `both_axes` argument

```{r}
mtcars |> 
  ggstackplot(
    x = mpg, y = c(wt, qsec),
    color = c("#E41A1C", "#377EB8"),
    both_axes = TRUE
  )

mtcars |> 
  ggstackplot(
    y = mpg, x = c(wt, qsec),
    color = c("#E41A1C", "#377EB8"),
    both_axes = TRUE
  )
```

## `alternate_axes` argument

```{r}
mtcars |> 
  ggstackplot(
    x = mpg, y = c(wt, qsec),
    color = c("#E41A1C", "#377EB8"),
    alternate_axes = FALSE
  )

mtcars |> 
  ggstackplot(
    y = mpg, x = c(wt, qsec),
    color = c("#E41A1C", "#377EB8"),
    alternate_axes = FALSE
  )
```

## `switch_axes` argument

```{r}
mtcars |> 
  ggstackplot(
    x = mpg, y = c(wt, qsec),
    color = c("#E41A1C", "#377EB8"),
    switch_axes = TRUE
  )

mtcars |> 
  ggstackplot(
    y = mpg, x = c(wt, qsec),
    color = c("#E41A1C", "#377EB8"),
    switch_axes = TRUE
  )

# and in combination with alternate_axes = FALSE
mtcars |> 
  ggstackplot(
    x = mpg, y = c(wt, qsec),
    color = c("#E41A1C", "#377EB8"),
    alternate_axes = FALSE,
    switch_axes = TRUE
  )

mtcars |> 
  ggstackplot(
    y = mpg, x = c(wt, qsec),
    color = c("#E41A1C", "#377EB8"),
    alternate_axes = FALSE,
    switch_axes = TRUE
  )
```

## `overlap` argument

```{r}
# define any overlap between 0 and 1
mtcars |> 
  ggstackplot(
    x = mpg, y = c(qsec, drat),
    color = c("#E41A1C", "#377EB8"),
    overlap = 0.3
  )

# full overlap
mtcars |> 
  ggstackplot(
    x = mpg, y = c(qsec, drat),
    color = c("#E41A1C", "#377EB8"),
    overlap = 1
  )
```

### different overlaps

```{r}
# different overlap between stack panels
mtcars |> 
  ggstackplot(
    x = mpg, 
    y = c(qsec, drat, wt, hp),
    color = c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3"),
    overlap = c(1, 0, 1)
  )

# and the horizontal version
mtcars |> 
  ggstackplot(
    y = mpg, 
    x = c(qsec, drat, wt, hp),
    color = c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3"),
    overlap = c(1, 0, 1)
  )
```

## `shared_axis_size` argument

The size of the shared axis often needs to be adjusted depending on which aspect ratio is intended. It is defined as fraction of a full panel.

```{r}
mtcars |> 
  ggstackplot(
    x = mpg, y = c(qsec, drat),
    color = c("#E41A1C", "#377EB8"),
    overlap = 1,
    # can be only 10% of a plot size as we're overlapping plots
    shared_axis_size = 0.1
  )
```

## `simplify_shared_axis` argument

Sometimes it's better to keep the shared axis on each panel.

```{r}
mtcars |> 
  ggstackplot(
    x = mpg, y = c(qsec, drat),
    color = c("#E41A1C", "#377EB8"),
    simplify_shared_axis = FALSE
  )

# also goes well with changing `both_axes`, `switch_axes` and/or `alternate_axes`
mtcars |> 
  ggstackplot(
    x = mpg, y = c(qsec, drat),
    color = c("#E41A1C", "#377EB8"),
    simplify_shared_axis = FALSE,
    alternate_axes = FALSE
  )
```

# The `template` argument

This is the most powerful argument. It defines which ggplot to use as the template for all plots in the stack. This can be an actual plot (just the data will be replaced) or a ggplot that doesn't have data associated yet. The possibilities are pretty much endless. Just make sure to always add the `theme_stacked_plot()` base theme (you can modify it more from there on). A few examples below:

## theme modifications

Add any modification to the overlying theme as you see fit.

```{r}
library(ggplot2)

# increase y axis text size
mtcars |>
  ggstackplot(
    x = mpg, y = c(qsec, drat),
    color = c("#E41A1C", "#377EB8"),
    template = 
      ggplot() + 
      geom_line() +
      theme_stackplot() +
      theme(
        axis.title.y = element_text(size = 20),
        axis.text.y = element_text(size = 16)
      )
  )

# increase the panel margins
mtcars |>
  ggstackplot(
    x = mpg, y = c(qsec, drat),
    color = c("#E41A1C", "#377EB8"),
    template = 
      ggplot() + 
      geom_line() +
      theme_stackplot() +
      theme(
        # increase left margin to 20% and top/bottom margins to 10%
        plot.margin = margin(l = 0.2, t = 0.1, b = 0.1, unit = "npc")
      )
  )
```

## custom geom data

```{r}
# use different geoms for different panels
# you can refer to y-stack panel variables with `.yvar` and x-stack panel variables with `.xvar`
mtcars |>
  ggstackplot(
    x = mpg, y = c(qsec, drat),
    color = c("#E41A1C", "#377EB8"),
    overlap = 1,
    template = ggplot() +
      geom_line(data = function(df) filter(df, .yvar == "qsec")) +
      geom_point(data = function(df) filter(df, .yvar == "drat")) +
      theme_stackplot()
  )
```

## additional plot elements

```{r}
# add some vertical bars
mtcars |>
  ggstackplot(
    x = mpg, y = c(qsec, drat),
    color = c("#E41A1C", "#377EB8"),
    overlap = 0.2, 
    template = 
      ggplot() + 
      geom_vline(xintercept = 20, linewidth = 4, color = "gray80") +
      geom_line() +
      theme_stackplot() 
  )
```

## axis modifications

```{r}
# add secondary x axis
mtcars |>
  ggstackplot(
    x = mpg, y = c(qsec, drat),
    color = c("#E41A1C", "#377EB8"),
    both_axes = TRUE,
    overlap = 0.1, 
    template = 
      ggplot() + 
      geom_line() +
      scale_x_continuous(
        # change axis name
        name = "this is my elaborate mpg axis",
        # this can be the same with dup_axis() or as here have a transformed axis
        sec.axis = sec_axis(
          trans = sqrt, 
          name = expression(sqrt(mpg)), 
          breaks = scales::pretty_breaks(5)
        )
      ) +
      theme_stackplot() 
  )
```

## additional aesthetics

```{r}
# add aesthetics to the plot
mtcars |>
  ggstackplot(
    x = mpg,  y = c(wt, qsec, drat),
    alternate_axes = FALSE,
    template = 
      ggplot() +
      aes(color = factor(cyl), linetype = factor(cyl), shape = factor(cyl)) +
      geom_line() +
      geom_point(size = 3) +
      theme_stackplot() 
  )
```


# The `add` argument

For even more specific plot refinements, the `add` argument provides an easy way to add ggplot components to specific panels in the stack plot. A few examples below:

## custom geoms

```{r}
# similar to the example `custom geom data`, the `add` argument can also
# be used to add specific geoms only to specific panels
mtcars |>
  ggstackplot(
    x = mpg, y = c(qsec, drat),
    color = c("#E41A1C", "#377EB8"),
    template = ggplot() + theme_stackplot(),
    add = list(
      # panel by name
      qsec = geom_line(), 
      drat = geom_rect(
        xmin = 20, xmax = 25, ymin = 3.2, ymax = 4.2, fill = "gray90") + 
        geom_point()
    )
  )
```

## custom themes

```{r}
# likewise custom themes can be added to individual panels with `add`
mtcars |>
  ggstackplot(
    x = mpg, y = c(qsec, drat),
    color = c("#E41A1C", "#377EB8"),
    template = ggplot() + geom_line() + theme_stackplot(),
    add = list(
      # panel by index
      geom_point() + theme(axis.title.y = element_text(size = 30))
    )
  )
```

## custom scales

```{r, message=FALSE}
# particularly useful is also the possibility to modify individual scales
mtcars |>
  ggstackplot(
    x = mpg, y = c(qsec, drat),
    color = c("#E41A1C", "#377EB8"),
    template = ggplot() + geom_line() + theme_stackplot(),
    add = list(
      # panel by index
      NULL,
      scale_y_continuous("$$ drat", labels = scales::label_dollar()) + 
        theme(axis.title.y = element_text(size = 30)) 
    )
  )
```

## legend positioning

```{r}
# another example of theme modification is the use of the `add` argument
# to specify legend positioning
mtcars |>
  ggstackplot(
    x = mpg,  y = c(wt, qsec, drat),
    color = c("#E41A1C", "#377EB8", "#4DAF4A"),
    template = 
      ggplot() + aes(linetype = factor(vs)) +
      geom_line() + theme_stackplot(),
    # switch legend position for middle plot
    add = list(qsec = theme(legend.position = "left"))
  )

mtcars |>
  ggstackplot(
    x = mpg,  y = c(wt, qsec, drat),
    color = c("#E41A1C", "#377EB8", "#4DAF4A"),
    template = 
      ggplot() + aes(linetype = factor(vs)) +
      geom_line() + theme_stackplot() +
      theme(legend.position = "none"),
    # re-include the middle panel legend on the plot
    # with some additional styling
    add = list(
      qsec = 
        theme(
          legend.position = c(0.2, 0.9),
          legend.title = element_text(size = 20),
          legend.text = element_text(size = 16),
          legend.background = element_rect(
            color = "black", fill = "gray90", linewidth = 0.5),
          legend.key = element_blank(),
          legend.direction = "horizontal"
        ) +
        labs(linetype = "VS")
    )
  )
```
